---
title: "ExtractUpstreamFeatures"
output: html_document
---

# Extract Features Upstream and Downstream of the TSS.


## 1. Load Libraries

```{r}
# load libraries
#library(txdbmaker)
library(GenomicFeatures)
library(Rsamtools)
library(Biostrings)
library(dplyr)
```

## 2. 

The indexed genome (.fna.fai), the genome, and the transcript database (created in create_TxDb.R) are first loaded.
```{r}
# set path to genome 
rpadi_genome = "./genome_gff/GCF_020882245.1_ASM2088224v1_genomic.fna"

# This will load the fai file according to the rpadi_genome file name using sprint.
rpadi_fafile = FaFile(rpadi_genome)

# set path to database (it should have info for chromosome starts and ends, see create_TxDb.R)
rpadi_txdb = loadDb("./genome_gff/rpadi_V4_txdb.sqlite")
```

## 3.

Now we will grab all the genes and their promoters, then extract regions up and downstream of the transcription start site.
```{r}
# extract genes 
rpadi_genes = genes(rpadi_txdb)

# view extracted genes
head(rpadi_genes)

# extract promoter regions around each gene TSS:
rpadi_promoters = promoters(rpadi_genes, upstream=1500, downstream=150)

# trim them so they don't exceed contig boundaries:
rpadi_promoters = trim(rpadi_promoters)

# now extract these (possibly shortened) promoter ranges
rpadi_upseqs = getSeq(rpadi_fafile, rpadi_promoters)

## if you want them oriented in the gene’s 5'->3' direction,
neg_idx = strand(rpadi_promoters) == "-"
rpadi_upseqs[neg_idx] = reverseComplement(rpadi_upseqs[neg_idx])

# rename the sequences using your gene IDs
names(rpadi_upseqs) = rpadi_genes$gene_id
```

## 4. 

For our dryR analysis we have a list of transcripts per each model that correspond to genes. These transcripts are in the "rna-XM" notation but our genes are in a different notation. Furthermore, one gene might be the parent of many transcripts.


This code will grab all of the promoters for each gene in our specific list. Then it will allow the user to save to directories.
```{r}

library(GenomicFeatures)
library(AnnotationDbi)
library(dplyr)
library(Biostrings)

# --- helpers ---
normalize_xm = function(x) {
  x = as.character(x)
  x = sub("^rna-", "", x)                 # drop "rna-" prefix
  x = sub("(\\.t\\d+)?(\\.\\d+)?$", "", x) # drop ".t{n}" and/or ".{version}" suffix
  x
}

# --- 1) Build XM to LOC mapping from TxDb (robust to column name variants) ---
tx_cols    = AnnotationDbi::columns(rpadi_txdb)
tx_field   = intersect(c("TXNAME","tx_name"), tx_cols)[1]
gene_field = intersect(c("GENEID","gene_id"), tx_cols)[1]
stopifnot(length(tx_field) == 1, length(gene_field) == 1)

tx_map_raw = AnnotationDbi::select(
  rpadi_txdb,
  keys    = AnnotationDbi::keys(rpadi_txdb, keytype = tx_field),
  keytype = tx_field,
  columns = c(tx_field, gene_field)
) %>% distinct()

# normalize mapping
tx_map = tx_map_raw %>%
  mutate(
    geneName   = .data[[tx_field]],
    gene_id    = as.character(.data[[gene_field]]),
    xm_norm    = normalize_xm(geneName),
    # ensure LOC prefix if the underlying gene_id is numeric/blank of LOC:
    loc_id     = ifelse(is.na(gene_id) | gene_id == "", NA_character_,
                        ifelse(grepl("^LOC", gene_id), gene_id, paste0("LOC", gene_id)))
  ) %>%
  filter(!is.na(xm_norm), !is.na(loc_id)) %>%
  select(xm_norm, loc_id) %>%
  distinct()

# quick sanity checks
cat("#TxDb map rows:", nrow(tx_map_raw), " = usable:", nrow(tx_map), "\n")

# --- 2) Writer that normalizes CSV XM ids, joins to LOC, and writes FASTA ---
write_promoters_for = function(df, out_fasta, id_col = "geneName") {
  stopifnot(id_col %in% colnames(df))
  csv_norm = df %>%
    transmute(xm_norm = normalize_xm(.data[[id_col]])) %>%
    distinct()

  # join to LOC via normalized XM
  mapped = csv_norm %>% left_join(tx_map, by = "xm_norm")

  # diagnostics
  n_csv     = nrow(csv_norm)
  n_mapped  = sum(!is.na(mapped$loc_id))
  loc_in_db = intersect(na.omit(mapped$loc_id), names(rpadi_upseqs))
  n_haveseq = length(loc_in_db)
  cat(sprintf(
    "[%s] CSV=%d | mapped XM→LOC=%d | with promoter seq=%d\n",
    out_fasta, n_csv, n_mapped, n_haveseq
  ))

  # write
  writeXStringSet(rpadi_upseqs[loc_in_db], out_fasta)
}

# --- 3) Write FASTAs (adjust paths as needed) ---
# example write_promoters_for(light_driven, "promoters_light_driven_1500bp_5to3.fasta")
# example write_promoters_for(endogenous,  "promoters_endogenous_1500bp_5to3.fasta")


```


## 5. Saving all dryR model promoter regions to directories for further analysis with the meme suite of tools.

Our dryR model transcript IDs can be found in the "./model_promoter_fastas/dryR_v2_geneIDs" directory.
Our promoter fasta files will ultimately be saved to the "./model_promoter_fastas/" directory. 
```{r}
# loading wheat models
wheat_model2 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/wheat_model2.txt", col.names = "geneName")
wheat_model3 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/wheat_model3.txt", col.names = "geneName")
wheat_model4 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/wheat_model4.txt", col.names = "geneName")
wheat_model5 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/wheat_model5.txt", col.names = "geneName")

# diet models
diet_model2 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/diet_model2.txt", col.names = "geneName")
diet_model3 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/diet_model3.txt", col.names = "geneName")
diet_model4 = read.table("./model_promoter_fastas/dryR_v2_geneIDs/diet_model4.txt", col.names = "geneName")

# write wheat promoter files
write_promoters_for(wheat_model2, "./model_promoter_fastas/wheat_model2_promoters_1500bp-150bp_5to3.fasta")
write_promoters_for(wheat_model3, "./model_promoter_fastas/wheat_model3_promoters_1500bp-150bp_5to3.fasta")
write_promoters_for(wheat_model4, "./model_promoter_fastas/wheat_model4_promoters_1500bp-150bp_5to3.fasta")
write_promoters_for(wheat_model5, "./model_promoter_fastas/wheat_model5_promoters_1500bp-150bp_5to3.fasta")

# write diet promoters
write_promoters_for(diet_model2, "./model_promoter_fastas/diet_model2_promoters_1500bp-150bp_5to3.fasta")
write_promoters_for(diet_model3, "./model_promoter_fastas/diet_model3_promoters_1500bp-150bp_5to3.fasta")
write_promoters_for(diet_model4, "./model_promoter_fastas/diet_model4_promoters_1500bp-150bp_5to3.fasta")
```

